## Background knowledge: R package / library

### not so precise definitions

A package is a collection of R functions, objects, compiled-code and data in a well defined format.

A library is a directory where R packages are stored

* Function: .libPaths()
* R_LIBS in [startup files](http://stat.ethz.ch/R-manual/R-devel/library/base/html/Startup.html) e.g. .Renviron / .RProfile

### Using R packages

To load a package from your library into the current session

* library()
* require()

Check the session info with

```{r sessionInfo}
library(stringr)
sessionInfo()
```


The difference between library() and require()

* echo in require() by default
     * require(stringr, quietly=TRUE)
* The actual difference. Try:

```{r difflibreq, echo=TRUE}
sucess1 <- library(MASS)
sucess2 <- require(MASS)
sucess1 # loaded packages in this session
sucess2 # boolean for success loading
```

Unload a package from the current session

* detach(package:MASS)

Installation of packages

* install.packages(devtools) ### from CRAN
* install.packages("~/devtools.tgz") ### from local file
* R CMD build --binary ~/devtools.tgz
* R CMD install ~/devtools.tgz

Uninstall

* remove.packages(fortune)
* R CMD REMOVE fortune


### Structure of an R package

tgz file = tar.gz (file compression)

```
./mypkg
|-- COPYING
|-- DESCRIPTION
|-- INDEX
|-- R/
|   |-- call_c.r
|   |-- call_f.r
|   `-- zzz.r
|-- README
|-- demo/
|   |-- 00Index
|   |-- RcallC.R
|   `-- RcallF.R
|-- man/
|   |-- test.c.Rd
|   `-- test.f.Rd
`-- src/
    |-- call_c.c
    `-- call_f.f
```

Absolute essential:

* R/
* DESCRIPTION file

Preferred

* man/
* NAMESPACE file

Optional

* data/
* inst/
* demo/
* tests/
* src/
* exec/
* po/
* NEWS file
* README
* INDEX

## R package development

### Styles

#### Oldskool style: [Writing R Extesions](http://cran.r-project.org/doc/manuals/R-exts.html)

DIY from scratch, Write documentation with Latex, etc.

#### Hadley ecosystem style: [Advanced R Programming](http://adv-r.had.co.nz/#package-development)

Automate all the tedious tasks with the devtools and related packages.

### Scope of this workshop

To package R files with R functions into an R package with documentation and tests using the Hadley ecosystem style.

Not cover

* /data, /po, /exec...
* packaging S3, S4, R5 objects
* packaging C / Fortran / Java code

### Development lifecycle

### create a package skeleton

#### require(devtools)

#### create("Dawkins)

Content created      | Purpose
---------------------|-----------------------------------------------------
R/Dawkins-package.r  | package documentation, namespace control
DESCRIPTION file     | metadata about the package
man/                 | documentation. In the Hadley Style: auto-generation

### Put weasel.R and monkey.R into Dawkins/R/

### Try to load all the functions in Dawkins/R/

#### load_all("./Dawkins")

With error messages complain the empty DESCRIPTION

### Edit the DESCRIPTION for the first time

NO BLANK LINE in the DESCRIPTION file!

Title: A package for Richard Dawkins' theory of gene centric view of evolution

Authors: youname <youremail@youremail.com>

Description: Functions to demostrate the difference between 

Junkyard Tornado and accumulative evolution

License: GPL-2

Add: Maintainer: yourname <yourname@yournam.com>

We will handle the problem of dependence later.

### Create a NAMESPACE for your package

every (good) package should have its own namespace.

The concept of namespace.

e.g. 

* require(MASS)
* MASS::glm.nb

You are calling the glm.nb function from the MASS namespace

External functions within the package namespace- for user to use without the packagename::function notation.

e.g.

* require(MASS)
* glm.nb ### because glm.nb is available from the external

Internal functions - can only be use within the package's namespace

Considerations for the NAMESPACE

* Which function should be available for the user as an external function?
      * You need to document all the external functions.
* Which package are the dependence?

For the Dawkins package:

* hurricane, evolution and experiment should be available as external functions, all the rest are helper functions.
* The dependence for Dawkins: ggplot2, plyr

### handling the dependence

R will install all dependence for you. Should specify in the DESCRIPTION.

The difference between Depends, Imports, Suggests

* Depends: works like require()
* Imports: make the external functions for the imported package usable in your package's namespace
* Suggests: Not absolute require but enhance the functionality

Use Imports whenever possible.

### Make ggplot2 and plyr as "Imports"

Edit the DESCRIPTION with the Imports field

Edit the package documentation file.

"#'"" is the special notation for use with the Roxygen2.

* @import ggplot2
* @import plyr
* can use @importFrom plyr ldply
    * if you only need a small number of function from a dependence.

run document("./Dawkins")

Have a look at the NAMESPACE and DESCRIPTION

remove the require() lines in the weasel.R file

Restart R session to make sure DESCRIPTION and NAMESPACE work

* require(devtools)
* load_all("./Dawkins")

Dependence seems to be working

### Documentation and make some functions external.

Using Roxygen2 to document. (Please check the original documentation)

Just above the evolution() function.

```{r}

#' Weasel program
#' 
#' This function demostrates the process of evolution as suggested by Dawkins.
#' 
#' @param origin the origin string
#' @param target the target string
#' @param numOffSpring Number of springs to reproduce
#' @param rateMutate Mutation Rate
#' @param maxGen Maximum number of generations
#' @examples
#' evolution()
#' @export

```

Take a look at man/ and NAMESPACE

Do the same for experiment and hurricane

```{r}


#' To find the optimal mutation rate
#' 
#' This function determines the optimal mutation rate for the evolution function
#' 
#' @param mutationRates A vector of mutation rate
#' @examples
#' experiment()
#' @seealso evolution
#' @export


#' The junkyard tornado simulation
#' 
#' This function demostrates the spontaneous generation of gene combination in the junkyard tornado metaphor.
#' 
#' @param origin the origin string
#' @param target the target string
#' @param maxGen Maximum number of generations
#' @examples
#' hurricane()
#' @seealso evolution
#' @export

```

As well as the package documentation file

```{r}
#' Dawkins
#' 
#' Doawkins provides functions to demostrate the concepts of gene centric view of evolution.
#'
```


### Try to build the package!

R CMD build ./Dawkins

### And check for problems

R CMD check ./Dawkins_0.1.tar.gz

### And install!

R CMD install ./Dawkins_0.1.tar.gz

### And use it!

* require(Dawkins)
* evolution()
* But you cannot use evolve()

### Congrats!

You have created your first package.

### Unit testing

Unit testing with test that (check the HKRUG #1 video demo)

all tests should be stored in inst/tests

and put a file called "run-all.R" inside tests/ with the content being

unit testing will be conducted when R CMD check.

```
library(testthat)
test_package("Dawkins")
```

### Brief intro to data

put the data into the data directory and document it in the package documentation under NULL if needed.

### Other useful techniques

* include a .Rbuildignore in your package directoty to exclude files during R CMD build

```
.git
.gitignore
.Rbuildignore
```
